\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-07-19 paper/ms.tex
page break

To start, we define a rank-$R$ semiseparable matrix as a matrix $K$ where the
elements can be written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.
In this representation, rather than storing the full $N \times N$ matrix $K$,
it is sufficient to only store the $N \times R$ matrices $U$ and $V$, and the
$N$ diagonal entries of $A$, for a total storage volume of $(2\,R+1)\,N$
floating point numbers.
Given a semiseparable matrix of this form, it is known that matrix-vector
products and matrix-matrix products can be evaluated in $\mathcal{O}(N)$
operations and the interested reader is directed to \citet{Vandebril:2007} for
more details about semiseparable matrices in general.

@@

\section{Examples}\sectlabel{examples}

@@

\subsection{Example 4: Asteroseismic oscillations}\sectlabel{astero}

@@

\subsection{Example 5: Exoplanet transit fitting}\sectlabel{transit}

@@

Table~\ref{tab:example-stats} lists the specifications of each example.
In this table, we list the computational cost required for a single likelihood
evaluation using a general Cholesky factorization\footnote{We use the Intel
Math Kernel Library \url{https://software.intel.com/en-us/intel-mkl}}
and the cost of computing the same likelihood using \celerite.
Combining this with the total number of model evaluations required to run the
MCMC algorithm to convergence (this number is also listed in
Table~\ref{tab:example-stats}), we can estimate the total CPU time required in
each case.
While the exact computational requirements for any problem also depend on the
choice of inference algorithm and the specific precision requirements, this
table provides a rough estimate of what to expect for projects at these
scales.
}

>>>>

To start, we define a rank-$R$ semiseparable matrix as a matrix $K$ where the
elements can be written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.
In this representation, rather than storing the full $N \times N$ matrix $K$,
it is sufficient to only store the $N \times R$ matrices $U$ and $V$, and the
$N$ diagonal entries of $A$, for a total storage volume of $(2\,R+1)\,N$
floating point numbers.
Given a semiseparable matrix of this form, the matrix-vector products and
matrix-matrix products can be evaluated in $\mathcal{O}(N)$ operations.
We include a summary of the key operations in \app{operations}, but the
interested reader is directed to \citet{Vandebril:2007} for more details about
semiseparable matrices in general.

@@

\newpage
\section{Examples}\sectlabel{examples}

@@

\newpage
\subsection{Example 4: Asteroseismic oscillations}\sectlabel{astero}

@@

\newpage
\subsection{Example 5: Exoplanet transit fitting}\sectlabel{transit}

@@

Table~\ref{tab:example-stats} lists the specifications of each example.
In this table, we list the computational cost required for a single likelihood
evaluation using a general Cholesky factorization and the cost of computing
the same likelihood using \celerite.
Combining this with the total number of model evaluations required to run the
MCMC algorithm to convergence (this number is also listed in
Table~\ref{tab:example-stats}), we can estimate the total CPU time required in
each case.
While the exact computational requirements for any problem also depend on the
choice of inference algorithm and the specific precision requirements, this
table provides a rough estimate of what to expect for projects at these
scales.
}
